Limited predictability of amino acid substitutions in seasonal influenza viruses

### 2021-01-13 article.tex

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
The trajectory of a mutation will depend on its own effect on fitness, the contribution of the genetic background on the same segment, and the effect of the remaining seven segments.
Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to the mutation.

@@

Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
In the extreme case were the two mutations always appear on the same strains, their two trajectories will be identical.
This ``nesting'' of frequency trajectories may introduce biases when performing statistical analysis.
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, it does not significantly change the results presented below.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{C} and for specific values of $\Delta t$. \textbf{C}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f0$.}
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, let's consider the long-term goal of predicting the probability that a mutation fixes in the population.
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen.
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency for A/H3N2 and A/H1N1pdm, respectively.
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}.
We also note that the fraction of lost trajectories increases sharply at small times with 40\% of mutations observed above 25\% frequency being lost within one year for A/H3N2, while it takes longer to fix a mutation in the whole population.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen.
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time.
The probability of a mutation fixing given that it is seen at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA.

@@

This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are observed.
In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

@@

Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions.
Mutations at these positions might mediate immune escape and might be expected to be under strong selection and show sweep-like behavior.
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}.
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes.
Only mutations at the 7 pivotal sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance.
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} show lower chances of fixation.
One should also note that many of these positions were determined post-hoc and might be enriched for positions that experienced rapid substitutions before the publication of the respective studies.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency, for A/H3N2 influenza. Figure \sref{fig:pfix_stratified_h1n1} shows the same analysis for A/H1N1. \textbf{A}: HA mutations with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: HA and NA mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: HA and NA mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

@@

The low power to predict frequency dynamics or fixation naturally triggers the question why the above methods have been found to work.
Picking representatives of the future and predicting frequency dynamics are distinct objectives and success at the former (as compared to random picks) is not necessarily inconsistent with a lack of predictable dynamics.
In fact, \citep{huddleston_integrating_2020} reports that the rate at which the frequency of a strain changes is often a poor predictor -- consistent with our observations here.
But despite the fact that future frequencies are not predicted by the LBI, the strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one.
While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, it does not seem to be predictive of influenza evolution because it measures fitness from genealogical structure.
Instead, we believe it picks closer than average strains simply because it has the tendency to be maximal at the base of large and dense clades.
These basal genotypes are closer to the future populations than the current tips of the tree and hence a better predictor on average.
The consensus sequence of all present strains performs slightly but consistently better than picking the strain with the highest LBI.
The consensus sequence is the best possible predictor for a neutrally evolving population, and does not attempt to model fitness in any way.

@@

\subsection*{Data and code availability} % (fold)
\label{sub:data}
	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}.
	Strain names and accession numbers are given as tables in two supplementary files. Outliers strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/tree/master/src/config} were removed. \\
	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}.
	% https://github.com/nextstrain/seasonal-flu/tree/081be2f487182252627418491be36cd01d645954/config
% subsection data (end)

@@

\subsection*{Measuring the geographical spread of a mutation} % (fold)
\label{sub:measuring_the_geographical_spread_of_a_mutation}
	For a mutation $X$ we define its regional distribution using the numbers $n_r(X)$ that represent the number of sequences sampled in region $r$ that carry $X$. Regional weights are then defined as
	$$ w_r(X) = \frac{n_r(X)}{\sum_{r}n_r(X)}.$$
	We can then measure the geographical spread $G(X)$ of $X$ by using the Shannon entropy of the probability distribution $w_r(X)$:
	$$G(X) = \sum_r w_r(X)\log(w_r(X)).$$
	$G(X)$ is a positive quantity that is larger when $X$ is equally present in many regions, and equal to zero when $X$ is concentrated in only one region.\\
	Region used are the ones defined in the \texttt{Nextstrain} tool~\cite{10.1093/bioinformatics/bty407}. Those are {North America}, {South America}, {Europe}, {China}, {Oceania}, {Southeast Asia}, {Japan \& Korea}, {South Asia}, {West Asia}, and {Africa}.
% subsection measuring_the_geographical_spread_of_a_mutation (end)

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/simulation_fitness.png}
		\caption{Fixation probability as a function of frequency for the simulations discussed in the main text, with trajectories stratified according to real fitness values. ``High'' and ``low'' fitness classes are defined with respect to the median value. \textbf{Left:} Simulation without antibodies and with changes to the fitness landscape every $dt=10$ generations. \textbf{Right:} Simulation with antibodies, with a new antibody every $dt=10$ generations. }
		\label{fig:simulations_fitness}
	\end{figure}

>>>>

The main underlying question asked in this work is the following: given a mutation $X$ in the genome of influenza that we observe at a frequency $f$ in the population at a given date, what can we say about the future of $X$?
The trajectory of a mutation will depend on its own effect on fitness, the contribution of the genetic background on the same segment, and the effect of the remaining seven segments.
Here, we investigate properties of broad categories of mutations effectively averaging over different genetic backgrounds to isolate the effects intrinsic to novel mutation in this category.

@@

Since genomic segments of influenza do not recombine, concurrent trajectories of mutations on the same segment will be correlated.
In the extreme case were the two mutations always appear in the same strains, their two trajectories will be identical.
This ``nesting'' of frequency trajectories may introduce biases when performing statistical analysis.
In section \ref{sub:correcting_for_nested_trajectories} of the SM, we investigate the effect of such nested trajectories by clustering them based on the strains that compose them.
While the number of trajectories that contribute to the analysis decreases with more aggressive clustering, it does not significantly change the results presented below.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel1.png}
	\caption{\textbf{A}: Sketch of the idea behind the short term prediction of frequency trajectories. Given a mutation that we have seen increasing in frequency and that we ``catch'' at frequency $f_0$ at time $t_0$, what can we say about the distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$? \textbf{B}: Distribution of future frequencies $P_{\Delta t}(f\vert\,f_0)$ for the trajectories shown in panel \textbf{C} and for specific values of $\Delta t$. \textbf{C}: All frequency trajectories of amino acid mutations in the A/H3N2 HA and NA genes that were absent in the past, are seen around $f_0=30\%$ frequency at time $t_0=0$, and are based on more than 10 sequences at each time point. Red curves represent mutations that will ultimately fix, blue the ones that will be lost, and black the ones for which we do not know the final status. Dashed horizontal lines (blue and red) represent loss and fixation thresholds. The thick black line is the average of all trajectories, counting those that fix (resp. disappear) as being at frequency 1 (resp. 0). Figure \sref{fig:trajectories_h3n2_varf0} shows equivalent figures for other values of $f_0$.}
	\label{fig:trajectories}
\end{figure*}

\subsection*{Prediction of fixation or loss}
Instead of predicting future frequency, we then considered the probability that a mutation fixes in the population.
We first estimate the fraction of frequency trajectories that either fix in the population or are lost, as well as the time it takes for one or the other to happen.
Panels \textbf{A} and \textbf{B} of figure~\ref{fig:fix_loss_active} shows the fraction of frequency trajectories in HA and NA that either have fixed, were lost or remained active as a function of the time elapsed since they were first seen above 25\% frequency for A/H3N2 and A/H1N1pdm, respectively.
Most mutations are either lost or become fixed after 2-3 years, with very few trajectories remaining active after 5 years.
This time scale of 2-3 years is consistent with the typical coalescence time observed in phylogenetic trees of A/H3N2 influenza \citep{rambaut_genomic_2008,yan_phylodynamic_2019}.
We also note that the fraction of lost trajectories increases sharply at small times with 40\% of mutations observed above 25\% frequency being lost within one year for A/H3N2, while it takes longer to fix a mutation in the whole population.

@@

We then examined the probability of mutations to fix in the population as a function of the frequency at which they are seen.
For different values of frequency $f$, we consider all trajectories that started at a null frequency and are seen in the interval $[f - 7.5\%, f + 7.5\%]$ at any given time.
The fixation probability of a mutation at frequency $f$, $P_{fix}(f)$, is then estimated by the fraction of those trajectories which terminate at a frequency larger than $95\%$, \emph{i.e.} our fixation threshold.
Panels \textbf{C} and \textbf{D} of figure~\ref{fig:fix_loss_active} show $P_{fix}(f)$ as a function of $f$ for NA and HA.

@@

%This dynamics is in apparent contradiction with evidence that influenza surface proteins in A/H3N2 are under strong selective pressure to evade human immune responses \citep{bhatt_genomic_2011}.
%If strong selection was present, we would expect rising amino acid mutations to fix at a distinctively higher frequency than the one at which they are observed.
%In an extreme case where most trajectories would be clean sweeps, $P_{fix}(f)$ should be close to $1$ for all but very small values of $f$.

@@

Next, we focused on previously reported antigenic sites in the A/H3N2 HA protein, referred to as \emph{epitope} positions.
Mutations at these positions might mediate immune escape and might be expected to be under strong selection and show sweep-like behavior.
We used four lists of relevant epitope positions from different sources comprising from 7 to 129 positions in the sequence of the HA1 protein~\cite{Koel976, luksza_predictive_2014, Shih6283, wolf_long_2006}.
Panel Fig.~\ref{fig:pfix_stratified}\textbf{B} shows fixation probability as a function of frequency for the four lists of epitopes.
Only mutations at the 7 pivotal sites reported in~\cite{Koel976} have higher chances of fixation than expected by chance.
No clear difference is found for the lists by \citet{luksza_predictive_2014, wolf_long_2006}, while positions from \citet{Shih6283} fix less often.
One should also note that many of these positions were determined post-hoc and might be enriched for positions that experienced rapid substitutions before the publication of the respective studies.

@@

\begin{figure*}
	\centering
	\includegraphics[width=0.9\textwidth]{./Figures/Panel3.png}
	\caption{Fixation probability $P_{fix}(f)$ as a function of frequency, for A/H3N2 influenza. Figure \sref{fig:pfix_stratified_h1n1} shows the same analysis for A/H1N1pdm. \textbf{A}: HA mutations with higher or lower LBI values, based on their position with respect to the median LBI value. \textbf{B}: Different lists of epitope positions in the HA protein. The authors and the number of positions is indicated in the legend. \textbf{C}: HA and NA mutations for binary positions, \emph{i.e.} positions for which we never see more than two amino acids in the same time bin. \textbf{D}: HA and NA mutations that appear once or more than once in the tree for a given time bin.}
	\label{fig:pfix_stratified}
\end{figure*}

@@

The low power to predict frequency dynamics or fixation naturally triggers the question why the above methods have been found to work.
Picking representatives of the future and predicting frequency dynamics are distinct objectives and success at the former (as compared to random picks) is not necessarily inconsistent with a lack of predictable dynamics.
In fact, \citet{huddleston_integrating_2020} reports that the rate at which the frequency of a strain changes is often a poor predictor -- consistent with our observations here.
But despite the fact that future frequencies are not predicted by the LBI, the strain with the highest LBI in the population is a better predictor of the future population than a randomly picked one.
While the LBI was shown to be a correlate of relative fitness and be predictive of fixation in mathematical models of evolution \citep{neher_predicting_2014}, it does not seem to be predictive of influenza evolution because it measures fitness from genealogical structure.
Instead, we believe it picks closer than average strains simply because it has the tendency to be maximal at the base of large and dense clades.
These basal genotypes are closer to the future populations than the current tips of the tree and hence a better predictor on average.
The consensus sequence of all present strains performs slightly but consistently better than picking the strain with the highest LBI.
The consensus sequence is the best possible predictor for a neutrally evolving population, and does not attempt to model fitness in any way.

@@

\subsection*{Data and code availability} % (fold)
\label{sub:data}
	The sequences used are obtained from the GISAID database~\cite{shu2017gisaid}.
	Acknowledgement tables with strain names, accession numbers, and originating and submitting labs are given as two supplementary files. Outlier strains listed at \url{https://github.com/PierreBarrat/FluPredictibility/tree/master/src/config} were removed. \\
	The code used to generate the figures presented here is available at \url{https://github.com/PierreBarrat/FluPredictibility}.
	% https://github.com/nextstrain/seasonal-flu/tree/081be2f487182252627418491be36cd01d645954/config
% subsection data (end)

@@

\subsection*{Measuring the geographical spread of a mutation} % (fold)
\label{sub:measuring_the_geographical_spread_of_a_mutation}
	For a mutation $X$ we define its regional distribution using the numbers $n_r(X)$ that represent the number of sequences sampled in region $r$ that carry $X$. Regional weights are then defined as
	$$ w_r(X) = \frac{n_r(X)}{\sum_{r}n_r(X)}.$$
	We can then measure the geographical spread $G(X)$ of $X$ by using the Shannon entropy of the probability distribution $w_r(X)$:
	$$G(X) = \sum_r w_r(X)\log(w_r(X)).$$
	$G(X)$ is a positive quantity that is larger when $X$ is equally present in many regions, and equal to zero when $X$ is concentrated in only one region.\\
	Region used are the ones defined in \texttt{Nextstrain/flu}~\cite{10.1093/bioinformatics/bty407}. Those are {North America}, {South America}, {Europe}, {China}, {Oceania}, {Southeast Asia}, {Japan \& Korea}, {South Asia}, {West Asia}, and {Africa}.
% subsection measuring_the_geographical_spread_of_a_mutation (end)

@@

	\begin{figure}
		\centering
		\includegraphics[width=0.9\textwidth]{SM_figures/simulation_fitness.png}
		\caption{Fixation probability as a function of frequency for the simulations discussed in the main text, with trajectories stratified according to real fitness values. ``High'' and ``low'' fitness classes are defined with respect to the median value. \textbf{Left:} Simulation with a purely additive landscape that changes every $\Delta t=10$ generations. \textbf{Right:} Simulation with antibodies, with a new antibody every $\Delta t=10$ generations. }
		\label{fig:simulations_fitness}
	\end{figure}
